Optimal strategies in mental rotation tasks

### 2013-05-14 man/final-project.tex
Write background section

% We do not know the form of $S$, and so cannot compute this integral
% analytically. We cannot rely on a simple Monte Carlo simulation to
% estimate it, either: because mental rotation is costly in terms of
% cognitive resources and must be performed in a sequence, we cannot
% evaluate $S(I_b|I_M)$ at arbitrary $I_M$. Instead, we must choose a
% relatively small, sequential set of $I_M$ to estimate the
% integral.

% Bayesian quadrature \cite{Diaconis:1988uo} provides an alternate
% method for evaluating the integral by placing a prior distribution on
% functions and then computing a posterior over values of the
% integral. While Bayesian quadrature itself does not address the issue
% of choosing points to evaluate the function at, recent work in
% statistics and machine learning has examined how to optimally select
% these samples \cite{Osborne:2012tm}.

\section{Computational model}

People are presented with two images, $X_a$ and $X_b$, which are the
coordinates of the vertices of two shapes (e.g., see Figure
\ref{fig:stimuli}). Participants must determine whether $X_a$ and
$X_b$ were generated from the same (albeit possibly transformed and
permuted) original shape, i.e., whether $\exists R,M\textrm{ s.t. }
X_b=MRX_a$, where $M$ is a permutation matrix and $R$ is a rotation
matrix.

>>>>

There is a long and rich history of work on mental rotation beginning
with \cite{Shepard1971}, who presented participants with pairs of
images such as those shown in Figure \ref{fig:mental-rotation}. These
pairs three-dimensional objects could either be: the same object
rotated in plane (Figure \ref{fig:mental-rotation}A), the same object
rotated in depth (Figure \ref{fig:mental-rotation}B), or two different
objects (Figure \ref{fig:mental-rotation}C; in this case, the
``different'' objects had reflective symmetry, but no rotational
symmetry). Participants had to determine whether the two images were
of the same object or not, and \cite{Shepard1971} famously found that
response times for plane and depth rotations had a strong linear
relationship with the minimum angle of rotation between the two
objects. The conclusion was that participants were visually
``rotating'' the objects in their minds.

For many years, this idea was contested, with some researchers arguing
the underlying cognitive processes were not visual in nature
(e.g. \cite{Pylyshyn1981}). In particular, \cite{Anderson1978} proved
that the mental imagery debate could not be resolved on the basis of
response time data alone. In recent years, however, a significant
amount of work has investigated the neural underpinnings of the mental
rotation phenomena, arguing that the results of brain imaging studies
support the existence of visual mental rotation \cite{Kosslyn1988,
  Kosslyn:2009tj}.

We assume that mental imagery is indeed a visual process, and turn to
the question of how, and when, it is used. Previous work has examined
how people might use imagery \cite{Hegarty2004, Schwartz1999} to solve
reasoning problems, and when they might use imagery as opposed to a
symbolic rule \cite{Schwartz:1996uy}. People seem to use imagery
methodically, which begs the question: what is the method?

Others studying people's pattern of behavior when engaged in
self-directed learning have found that they tend to use a strategy of
\textit{active learning} \cite{Gureckis:2012gu, Markant:2012uu}. When
given a choice as to the information they can obtain, people will not
choose the information randomly (as is the case with a \naive{} Monte
Carlo sampler), but according to some utility function such as
information gain \cite{Nelson2007, Markant:2012uu}. We propose that
such a strategy can be combined with the notion of mental imagery as a
tool: an optimal model of mental rotation may, perhaps, rely on an
active strategy. In the next sections, we investigate this idea with
several different approaches.

\section{Computational-level model}

People are presented with two images, $X_a$ and $X_b$, which are the
coordinates of the vertices of two-dimensional shapes similar to those
used by \cite{Cooper:1975wp} (e.g., see Figure
\ref{fig:stimuli}). Participants must determine whether $X_a$ and
$X_b$ were generated from the same (albeit possibly transformed and
permuted) original shape, i.e., whether $\exists R,M\textrm{ s.t. }
X_b=MRX_a$, where $M$ is a permutation matrix and $R$ is a rotation
matrix.
